---
title: "Kaggle Titanic"
author: "bweigel"
date: "April 29, 2016"
output:
  knitrBootstrap::bootstrap_document:
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

# So Kaggle ... #

This is my first attempt at Kaggle. I have registered some time ago, but never rwally got around to doing an y work. Mainly because I cannot really motivate to do something that seems more like a chore, than fun. Anyway, when I first looked at the Titanic dataset I though: "This should be straight-forward." And since I had done some basic data sciency stuff and I had also taken the course of Trevor Hastie and ? Tibshiriana I just couldn't motivate myself to actially start working.
So here we go.

```{r knitrOptions, cache=FALSE}
library(knitr)
opts_chunk$set(message=F, error=T, warning = F, echo=F, dev='CairoSVG')
```

```{r prerequesites}
library(dplyr)
library(magrittr)
library(pROC)
library(boot)
library(glmnet)
library(data.table)
library(ggplot2)
# function for plotting barplots according to results of an aggreate function
# input is the same as aggregate()
# returns a function, where the axis labels can be changed through the parameters xlab, ylab, main
barplot0 <- function(...){
  df <- aggregate(...)
  return(function(main=NULL, xlab=NULL, ylab=NULL){
    xlab <- ifelse(is.null(xlab), names(df)[1], xlab)
    ylab <- ifelse(is.null(ylab), names(df)[2], ylab)
    main <- ifelse(is.null(main), NA, main)
    barplot(height = df[,2], names.arg = df[,1]%>%as.character(),  xlab = xlab, ylab=ylab)
  })
}


```

# The dataset #

Loading the dataset at first is a good idea. 
There are 13 columns in the data: 

 - PassengerID:   Just a unique number                                                [Int]
 - Survived:      did the passanger survive? yes:1/no:0                               (boolean)
 - PClass:        Passenger Class (1 - 3)                                             (categorical)
 - Name:          Passenger Name
 - Sex:           male/female                                                         (categorical)
 - Age:           Passenger Age                                                       [float] (should be int, but estimated ages are in xx.5 form)
 - SibSp:         Number of siblings/spouses aboard                                   [int]
 - ParchA:        Number of parents/children aboard                                   [int]
 - Ticket:        Ticket number                                                       [string]
 - Fare:          Passenger Fare                                                      [float]
 - Cabin:         Number of the cabin the passener was in                             [string]
 - and Embarked:  Port of Embarkation (C = Cherbourg; Q = Queenstown; S = Southampton) [categorical]
 
 
## Descriptive Statistics on the dataset ##
 
 A quick `summary()` of the dataframe gives some hints about the dataset.
 
```{r load-dataset}
df <- read.csv("data/train.csv", stringsAsFactors = F, header=T, na.strings = "")
summary(df)
```

The mean of Survived shows at first sight, that from this set over 60% of the passengers perished in the cold north atlantic waters. Only 38.4% of the passengers survived. For convenience I will set the Survived column to the corresponding logical value (`TRUE` or `FALSE`).

```{r}
# treat categorical variables as such
df <- df %>% mutate(Survived = as.logical(Survived),
                    Pclass = factor(Pclass, levels = 1:3, labels = c("1st", "2nd", "3rd")),
                    Sex = factor(Sex, levels = unique(Sex), labels = unique(Sex)),
                    Embarked = factor(Embarked, levels = c("S", "C", "Q"), labels = c("S. Hampton", "Cherbourg", "Queenstown")))

#drop cabin column because of many NAs
df <- df %>% select(-Cabin)
#save file
save(df, file="titanic.Rda")

```

Some (not-so-)fun facts about the data (in text format):

```{r, echo=F, eval=FALSE}
aggregate(Survived ~ Sex, df, mean)  # how much of the passengers of each class sußrvived?
aggregate(Survived ~ Pclass, df, mean)  # how much of the passengers of each class sußrvived?
aggregate(Age ~ Pclass, df, mean)  # average age in each class
aggregate(Survived ~ Embarked, df, mean)  # how much of the passengers vurvived depending on the city of embarkemnt?
aggregate(Sex ~ Pclass, df, function(x){ifelse(x == "male", 1, 0) %>% mean()})  # amount of males in each class 

aggregate(Fare ~ Survived, df, mean)    # how much fare did the average survivor pay?
```

```{r descriptive, echo=F, fig.height=6, fig.width=8}
par(mfrow=c(2,2))
barplot0(Age ~ Survived, df, mean)()
barplot0(Survived ~ Sex, df, mean)()
barplot0(Survived ~ Embarked, df, mean)()
barplot0(Survived ~ Pclass, df, mean)(xlab="Passenger Class")
barplot0(Age ~ Pclass, df, mean)(ylab="Avg. Age", xlab="Passenger Class")
barplot0(Sex ~ Pclass, df, function(x){ifelse(x == "male", 1, 0) %>% mean()})(ylab="male passengers", xlab="Passenger Class")
barplot0(Fare ~Survived, df, mean)()
barplot0(Survived ~ Embarked, df, mean)()
```

## Missing values ##

There are 687 NAs in _Cabin_, 177 in _Age_ and 2 in _Embarked_. Since so much data is missing from the Cabin column it is removed. `r is.na(df) %>% colSums`.

```{r removeCabin}
df <- select(df, -Cabin)
```

The Age column raises some concern, since there are 177 missing values (`NA`). So lets first take a look at all the rows, where there is no age.
 
```{r}
df[which(is.na(df$Age)),] %>% summary
# Full dataset
df %>% summarise(Survived=mean(Survived),
                                        percent.male=mean(ifelse(Sex=="male",1,0)),
                                        Pclass = mean(Pclass),
                                        Fare = mean(Fare),
                                        SibSp = mean(SibSp),
                                        Parch = mean(Parch))
# Only passengers without age
df[which(is.na(df$Age)),] %>% summarise(Survived=mean(Survived),
                                        percent.male=mean(ifelse(Sex=="male",1,0)),
                                        Pclass = mean(Pclass),
                                        Fare = mean(Fare),
                                        SibSp = mean(SibSp),
                                        Parch = mean(Parch))
```

The mortality of the passengers without an age is 10% higher than the average mortality (71% vs 62%). Also it seems that passengers w/o age were mostly in the lower classes 2 and 3 (avg class 2.6), male (70%) with rather cheap tickets (22 pounds) and were travelling without parents or children.

# Simple logistic regression #

At first I'll try a simple logistic regression and leave out any rows that contain `NA` values. Since the cabin column has 687 `NA` values this column was dropped before.

```{r}
#remove rows with NA values
dfWork <- df[!(is.na(df) %>% rowSums() %>% as.logical), ]

# split into train and test set
samples <- sample(nrow(dfWork), size = 0.6*nrow(dfWork))
dfWork_train <- dfWork[samples,]
dfWork_test <- dfWork[-samples,]

# perform simple logistic regression
logMod1 <- glm(Survived ~ ., family = binomial(link = "logit"), data = dfWork_train)
```

Unsurprisingly this simple approach doesn't work. The algorithm doesn't converge.
Let's fiddle around with the formula a bit.

```{r fitLogMod}
set.seed(1337)  
# simpler model
logMod1 <- glm(Survived ~ Pclass + Age + Sex + SibSp + Parch + Fare + SibSp:Age + SibSp:Pclass, family = binomial(link = "logit"), data = dfWork_train)

# seems like Parch and Fare are not significant as parameters for prediction
summary(logMod1)

# simplify model using step()
logMod2 <- step(logMod1)
summary(logMod2)
# as expected Parch and Fare are dropped
```

Now let's check the performance of our new model on te test set.

```{r predictLogMod}
# predict outcome
newY <- predict(logMod2, newdata = dfWork_test, type = "response")
dfAnalyze <- cbind(dfWork_test, SurvPred=newY) # this is for plotting later on
# compare prediction and actual values in confudion table
newY <- ifelse(newY <= 0.5, F, T)
confMat <- table(dfWork_test$Survived, newY)
knitr::kable(confMat)
confMat <- confMat/sum(confMat)
knitr::kable(confMat/sum(confMat))
```

This simple logistic regression model already does a reasonable well job of predicting the right outcome.
It classifies `r round((diag(confMat) %>% sum)*100,  2) `% of all cases into the right group.

```{r plotResultsLogMod}
ggplot(dfAnalyze %>% filter(SibSp == 0), aes(x = Age, y = SurvPred, color=Pclass, shape=Sex)) + 
  geom_point()  +
  geom_line() +
  labs(y="Predicted probability of Survival")
```


```{r glmCV}
logMod1 <- glm(Survived ~ Sex + Age + Embarked + Fare + Pclass + Parch + SibSp, dfWork, family= binomial(link="logit"))
logMod1 <- step(logMod1)
save(logMod1, file="models/logMod.Rda")
```

The 5-fold cross validation error or of the model `r print(logMod1$formula)` is approximately `r cv.glm(dfWork, logMod1, K=5)$delta[1]`.
